---
title: "Assignment 2: Physical Properties of Lakes"
author: "Yikai Jing"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Water Data Analytics on the physical properties of lakes. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After completing your assignment, fill out the assignment completion survey in Sakai. 

Having trouble? See the assignment's answer key if you need a hint. Please try to complete the assignment without the key as much as possible - this is where the learning happens!

Target due date: 2022-01-25

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, lubridate, and rLakeAnalyzer packages
3. Import the NTL-LTER physical lake dataset and set the date column to the date format. 
4. Using the `mutate` function, add a column called Month. Remove temperature NAs.
5. Set your ggplot theme (can be theme_classic or something else)

```{r}
#Verify your working directory
getwd()

# load packages
library(tidyverse)
library(lubridate)
library(rLakeAnalyzer)
library(dataRetrieval)
library(cowplot)

# Load data
NTLdata <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
#set the date column to the date format
NTLdata$sampledate <- as.Date(NTLdata$sampledate, format = "%m/%d/%y")
#add a column called Month
NTLdata <- NTLdata %>%
  mutate(Month = month(sampledate))
#Remove temperature NAs
NTLdata <- NTLdata %>%
  drop_na(temperature_C)

# set ggplot theme
theme_set(theme_classic())
```

## Creating and analyzing lake temperature profiles

5. For the year 1993, plot temperature and dissolved oxygen profiles for all six lakes in the dataset (as two separate ggplots). Use the `facet_wrap` function to plot each lake as a separate panel in the plot. Plot day of year as your color aesthetic and use a reverse y scale to represent depth. 

What seasonal trends do you observe, and do these manifest differently in each lake?

```{r}
#?facet_wrap
#?vars

NTLdata1993 <- NTLdata %>%
  filter(year4 == "1993")

TempPlot <-
ggplot(NTLdata1993, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = expression("Temperature "(degree*C)), y = "Depth (m)", color = "DOY")+
  facet_wrap(vars(lakename))+
  ggtitle("Temperature of Lakes in 1993") +
  theme(plot.title = element_text(hjust = 0.5))
TempPlot

DOPlot <-
ggplot(NTLdata1993, aes(x = dissolvedOxygen, y = depth, color = daynum)) +
  geom_point(alpha = 0.5) +
  scale_y_reverse() +
  scale_color_viridis_c() +
  labs(x = "Dissolved Oxygen (mg/L)", y = "Depth (m)", color = "DOY")+
  facet_wrap(vars(lakename))+
  ggtitle("Dissolved Oxygen of Lakes in 1993") +
  theme(plot.title = element_text(hjust = 0.5))
DOPlot

# From May to September, the temperature rises slowly. And from October to November, the temperature drops rapidly. Except Central Long Lake and TUesday Lake, all the 4 lakes heve the frozen period in December.
# From May to September, DO decreases slowly. And from October to November, DO increased rapidly.
# dimictic: East Long Lake, Peter Lake, West Long Lake
# monomictic: Paul Lake, Tuesday Lake
# polymictic: Central Long Lake
# amictic: 
```

6. Create a new dataset that calculates thermocline depths for all lakes on all dates (hint: you will need group by lake, year, month, DOY, and sample date).
7. Plot thermocline depth by day of year for your newly made dataset. Color each point by lake name, make the points 50% transparent, and choose a color palette other than the ggplot default. 
8. Create a boxplot of thermocline depth distributions split up by lake name on the x axis and by month as the fill color (hint: you will need to set Month as a factor). Choose a color palette other than the ggplot default, relabel axes and legend, and place the legend on the top of the graph.

Do all the lakes have a similar seasonal progression of thermocline deepening? Which lakes have the deepest thermoclines, and how does this relate to their maximum depth? 

```{r}
NTLdata_thermo <- NTLdata %>%
group_by(lakename, daynum, Month, year4, sampledate) %>%
summarise(thermocline = thermo.depth(wtr = temperature_C, depths = depth, seasonal = FALSE)) %>%
filter(thermocline > 1)
?thermo.depth

thermoclineplot <-
  ggplot(NTLdata_thermo, aes(x = daynum, y = thermocline, color = lakename)) +
  geom_point(alpha = 0.5) + #alpha = 0.5 : make the points 50% transparent
  scale_color_viridis_d()+
  labs(x = "DOY", y = "Thermocline depth (m)", color = "Lake")
thermoclineplot
#?scale_color_viridis_d()

thermoclineplot2 <-
  ggplot(NTLdata_thermo, aes(y = thermocline, x = lakename, fill = as.factor(Month))) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(x = "Lake", y = "Thermocline depth (m)", fill = "Month") +
  theme(legend.position = "top")
thermoclineplot2

depthplot <-
  ggplot(NTLdata, aes(y = depth, x = lakename, fill = as.factor(Month))) +
  geom_boxplot() +
  scale_fill_viridis_d() +
  labs(x = "Lake", y = "Depth (m)", fill = "Month") +
  theme(legend.position = "top")
depthplot

CombinedPlot <- 
  plot_grid(thermoclineplot2, depthplot,
            ncol = 1)
CombinedPlot
#Yes, their thermocline deepening all increase with time.
#Peter Lake has the deepest thermoclines, this as positively correlated with their maximum depth.
```

